Online Variable Elimination for Credal Polytrees

### 2017-03-02 basics.tex
More on basics

\section{Basics}\label{sec:basics}
\subsection{Bayesian and credal networks}
Given a discrete variable $X$, $\mathcal{X}$ denotes the set of its possible values and $x$ a generic element of this set. $P(X)$ is a probability mass function over $X$ and $P(x)$ the probability assigned to the singleton $x\in\mathcal{X}$. $K(X)$ denotes instead a convex set of probability mass functions over $X$, which is also called \emph{credal set} \cite{xxx}. The convex hull operator is denotes as $\mathrm{CH}$ (thus as CSs are convex $K(X)=CH[K(X)]$). In particular we focus on finitely generated credal sets characterised whose number of extreme points is therefore finite. Notation $\mathrm{ext}[K(X)]$ is used to denote this finite set of mass functions.

Given a set of variables $\bm{X}:=(X_1,\ldots,X_n)$, a BN is a pair composed by a directed acyclic graph $\mathcal{G}$ whose nodes are in one-to-one correspondence with the elements of $\bm{X}$ and a collection of conditional mass functions $P(X_i|X_{\pi_i})$ over $X_i$, one for each $x_{\pi_i}\in\mathcal{X}_{\Pi_i}$. In a credal network the collection of conditional mass functions is simply replaced by a credal set $K(X_i|\pi_i)$. While a BN defines a joint mass function $P(\bm{X})$ which factorises as follows
\[ P(\bm{x}) = \prod_i P(x_i|\pi_i) \]
a credal net defines a joint CS $K(\bm{X})$ whose extreme points factorised as in xxx, with the specification. Say that this is the strong independence case and cite some paper by Jasper to mention that other indep have been proposed.

It might be worth to mention that a CN is equivalent to (an exponential number) of CNs.

\subsection{Updating and Variable elimination}
Let us consider a Bayesian network over $\bm{X}$. Without lack of generality we can assume that the variable of interest $X_0$. An observation $x_E$ of a set of variables $X_E \subseteq \bm{X} \setminus X_0$ such that $P(x_E)>0$, updating is intended as the computation of:
\[ P(x_0|x_E) =\frac{\sum_i \prod_i }{\sum_i \prod_i} \]
where the right-hand side simply follows from the factorisation in \cite{x} the notion marginalisation of the unqueried and unobserved variable and the Bayes' theorem. For Bayesian networks updating is intended the identification of the lower/upper bounds.

>>>>

\section{Basics}\label{sec:basics}
\subsection{Bayesian and credal networks (BNs \& CNs)}
Given a discrete variable $X$, $\mathcal{X}$ denotes the set of its possible values and $x$ a generic element of this set. $P(X)$ is a probability mass function (PMF) over $X$ and $P(x)$ the probability assigned to the atom $x\in\mathcal{X}$. $K(X)$ denotes instead a set of PMFs over $X$, which is also called \emph{credal set} (CS). The bounds of the expectation of a linear function of $X$ w.r.t. $K(X)$ are not affected by the removal of the inner points of $K(X)$. We denote as $\overline{K}(X)$ the result of this operation. Here we focus on CSs with a finite number of extreme (i.e., non inner) points.

Given a set of variables $\bm{X}:=(X_0,X_1,\ldots,X_n)$, a BN is a pair composed by a directed acyclic graph $\mathcal{G}$ whose nodes are in one-to-one correspondence with the elements of $\bm{X}$ and a collection of PMFs $P(X_i|x_{\pi_i})$, one for each value of $x_{\pi_i}$, for each $X_i \in \bm{X}$, with $\Pi_i$ set of variables corresponding to the parents (i.e., predecessors according to $\mathcal{G}$) of $X_i$. In a CN specification, each $P(X_i|x_{\pi_i})$ is simply replaced by a CS $K(X_i|x_{\pi_i})$.

A BN defines a joint PMF $P(\bm{X})$, which factorises as \begin{equation}\label{eq:joint}
P(\bm{x}) = \prod_{i=0}^n P(x_i|\pi_i),
\end{equation}
for each $\bm{x}\in\mathcal{\bm{X}}$ and where the values of $x_i$ and $x_{\pi_i}$ are those consistent with $\bm{x}$. Similarly, a CN defines a joint CS $K(\bm{X})$ whose elements are PMFs obtained as in Eq.~\ref{eq:joint} with $P(X_i|\pi_i)\in K(X_i|\pi_i)$.

\subsection{Standard, Credal, and Bound Udating}

Consider a BN and a CN over $\bm{X}$. Without lack of generality we can assume that the variable of interest $X_0$. Given a set of variables $X_E$, denotes as $x_E\in\mathcal{X}_E$ an instance of these variables. Updating a BN is intended as the computation of $P(x_0,x_E)$, obtained by summing out the variables in $\bm{X}\setminus \{X_0,X_E\}$ from Eq.~\ref{eq:joint}, for each $x_0\in\mathcal{X}_0$, from which $P(X_0|x_E)$ can simply obtained by normalisation, provided that $P(x_E)>0$.

We can analogously intend updating in credal networks as the computation of the set $K(X_0,x_E)$, from which, by elementwise normalisation we obtain the updated CS $K(X_0|x_E)$. We call this CN inference task credal updating (CU) and distinguish it from the computation of the bounds $\underline{P}(x_0|x_E)$ and $\overline{P}(x_0|x_E)$ which is called here \emph{bound updating}.

\subsection{Variable Elimination (VE) in BNs}

VE is a standard approach to PU in BNs. The idea is to perform marginalisation only over the combination of local models including the variable to eliminate.

Given two (possibly joint )variables $X_I$ and $X_J$, a conditional probability table (CPT) $P(X_I|X_J)$ is a collection of PMF over $X_I$, one for each $X_J\in\mathcal{X}_J$. Given twp CPTs $P(X_I|X_J)$ and $P(X_K|X_L)$ we can simply obtain a new CPT, denoted as $P(X_I|X_J)\otimes P(X_K|X_L)$ by combination. The combined CPT is defined over $X_I \cup X_L$ given $X_J \cup X_M \setminus \{ X_I \cup X_J\}$. Given a CPT $P(X_I,X_J|X_K)$
the marginalization of $X_J$ is denoted as $\sum_{X_j} P)(X_I|X_K)$ being simply a CPT $P(X_I|X_K)$ obtained by summing out $X_J$ on each conditional PMF. Finally, given a CPT $P(X_I,X_J|X_K)$ and a state $x_J\in\mathcal{X}_J$ we cal focusing a CPT $P(X_I,x_j|X_K)$. Note that strictly speaking this is not a CPT over $X_I$ because of the non-normalisation w.r.t. $X_I$.

With this simple algebraic structure (see \cite{kohlas2003} for a xxx), we formulate the VE algorithm as in Table X.

\begin{algorithm}
	\caption{VE algorithm
		%\vskip 0.6mm [Parameters] $s$ (maximum number of no-improve iterations) and $t$ (number of restarts) 
		\vskip 0.6mm [INPUT] a BN specification, an evidence $(x_{n-n_e},\ldots,x_n)$%\\credal network specification $\{ K(X_i|\pi_i) \}_{i=0,\ldots,n}^{\pi_i\in\Omega_{\Pi_i}}$
		\vskip 0.6mm [OUTPUT] $K(X_o|x_{n-n_e},\ldots,x_n)$} %an upper approximation of $\underline{P}(x_0)$\label{algo:glp}}
	\begin{algorithmic}[1]
		%\State $pp \gets 1.0$
		\State $\Phi := \{ P(X_i|\Pi_i) \}_{i=0}^n$ 
		\For{$k \gets n,\ldots,0$}%\Comment{random restarts}
		\State $\phi_k =1$ %$\tilde{P}(X_i|\pi_i)$ $\gets$ randomly pick from $\mathrm{ext}[K(X_i|\pi_i)]$ $\forall i,\pi_i$ \Comment{initialization}
		\For{$\phi \in \Phi$}
		\If{$X_k$ is in the argument of $\phi$}
		\State $\phi_k \gets \phi_k \otimes \phi$
		\State $\Phi \gets \Phi \setminus \{ \phi \}$
		\EndIf
		\EndFor
		\EndFor
		\If{$k = 0 | E$}
		\State $\phi_k \gets \sum_{X_k} \phi_k $
		\EndIf
		\State $\Phi \gets \Phi \cup \{ \sum_{X_k} \phi_k \}$
		\State {\bf return} $pp$ 
	\end{algorithmic}
\end{algorithm}

To perform PU in a BN
here the right-hand side simply follows from the factorisation in \cite{x} the notion marginalisation of the unqueried and unobserved variable and the Bayes' theorem. For Bayesian networks updating is intended the identification of the lower/upper bounds.

### 2017-03-02 intro.tex
More on basics

\section{Introduction}\label{sec:intro}
Bayesian networks \citep[BNs]{2009:koller} are the most popular class of multivariate probabilistic models based on directed graphs. These tools are subject of intense theoretical investigation and their applications are ubiquitous, e.g., \cite{landuyt2013,de2013}. A common inference task with BNs, called \emph{updating}, consists in the computation of the probabilities of a variable of interest possibly given an instantiation of some other variables. BNs updating is a NP-hard task \citep{cooper1990}, which might become tractable under some additional assumption. If the standard \emph{variable elimination} \citep[VE]{kohlas} scheme is adopted, both time and space complexity are exponential in the \emph{treewidth}, which roughly describes how much a directed graph is similar to a singly-connected model (also called a polytree). Efficient updating can be therefore achieved (among others) in polytree-shaped BNs. This is basically equivalent to Pearl's message-propagation in polytrees \cite{pearl}.

Credal networks \citep[CNs]{cozman2000a} are a generalization of BNs based on the imprecise probability theory \citep{walley1991}. As well as a BN is specified by \emph{local} (w.r.t to the graph) probability mass functions, in a CN these are replaced by convex sets of them. We consequently call \emph{credal set updating} the task of  computing the credal set made of all the posterior mass functions for the queried variable given the evidence computed for all the BNs consistent with the CN specification. Here we call this task credal-set updating, while most of the authors intend updating as the computation of the lower and upper bounds of the posterior probability with respect to the CN specification. To avoid confusion, we call this latter task \emph{probabilistic updating} (PU).

Being generalisations of their Bayesian counterpart, both tasks are NP-hard. Yet, unlike the case of BNs, PU remains NP-hard even in polytree-shaped CNs \citep{maua2014}, while an efficient scheme has been developed only for the special case of polytrees with all the variables binary \citep{fagiuoli1998a}. Note that for binary queries PU and CU coincides.\footnote{A credal set over a binary variable cannot have more thant two xxx.}

Most of the literature about inference in CN is focused on approximate procedures for PU \cite{xxx}. Two remarkable exception are the early attempt of \cite{rocha} and a more recent paper by \cite{maua} where CU by VE is considered. In the first case, the convexification step, which is crucial to keep a bound on the exponential explosion on the number of extreme points, is performed. Unfortunately, as we show later, this induces an outer approximation. To avoid the outer approximation and keep the inferences exact, the second paper assumes the models extensively specified. In the present paper, we stick on the second approach, but assuming that the original model is separately specified. We therefore develop a number of online procedure which allow to perform the algebraic operations required by inference without xxx. Moreover, we suggest to perform convexification only on the space of single variables xxx.

>>>>

\section{Introduction}\label{sec:intro}
Bayesian networks \citep[BNs,][]{2009:koller} are the most popular class of multivariate probabilistic models based on directed graphs. These tools are subject of intense theoretical investigation and their applications ubiquitous, \citep[e.g.,][]{landuyt2013,de2013}. A common BN inference task, called \emph{updating}, consists in the computation of the probabilities of a queried variable given an instantiation of some other variables. BNs updating is a NP-hard task \citep{cooper1990}, which might become tractable under some additional assumption. Given an ordering of the variables, the \emph{variable elimination} \citep[VE,][]{zhang1994simple} is a standard procedure to update BNs. Yet, VE complexity depends on the ordering, and finding the optimal order is in its turn NP-hard. This is not the case with singly connected topologies (also called a polytrees, see, e.g., Fig.~\ref{fig:polytree}), for which an optimal ordering, which takes complexity exponential in the maximum number of parents, can be easily found in linear time.%\todo[inline]{Mention Pearl's algorithm too?}

Credal networks \citep[CNs,][]{cozman2000a} are a generalization of BNs based on the imprecise probability theory \citep{walley1991}. While a BN is specified by local (w.r.t. the graph) probability mass functions,these are replaced by sets of mass functions (also called \emph{credal} sets) in CNs. We call \emph{credal set updating} (CU) the task of computing the credal set of all the mass functions for the queried variable given the evidence computed in all the BNs consistent with the CN specification (i.e., with the local credal sets). Most of the authors intend CN updating as the computation of the lower and upper bounds of the posterior probability with respect to the CN specification. We call this latter task \emph{probabilistic updating} (PU). Being generalisations of their Bayesian counterpart, both tasks are NP-hard. Yet, unlike the case of BNs, PU remains NP-hard even in polytree-shaped CNs \citep{maua2014}, while an efficient exact scheme has been developed only for the special case of binary polytrees \citep{fagiuoli1998a}, for which CU and PU coincides.

\emph{*** This last part of the section has to be finished ***} Most of the literature about inference in CN is focused on approximate procedures for PU \cite{xxx}. Two remarkable exception are the early attempt of \cite{rocha} and a more recent paper by \cite{maua} where CU by VE is considered. In the first case, the convexification step, which is crucial to keep a bound on the exponential explosion on the number of extreme points, is performed. Unfortunately, as we show later, this induces an outer approximation. To avoid the outer approximation and keep the inferences exact, the second paper assumes the models extensively specified. In the present paper, we stick on the second approach, but assuming that the original model is separately specified. We therefore develop a number of online procedure which allow to perform the algebraic operations required by inference without xxx. Moreover, we suggest to perform convexification only on the space of single variables xxx.

The paper is organised as follows.
